---
title: "R Notebook"
output: html_notebook
---

Code: Scatterplot of the relationship between HRs and wins

```{r}
library(Lahman)
library(tidyverse)
library(dslabs)
ds_theme_set()

Teams %>% filter(yearID %in% 1961:2001) %>%
    mutate(HR_per_game = HR / G, R_per_game = R / G) %>%
    ggplot(aes(HR_per_game, R_per_game)) + 
    geom_point(alpha = 0.5)
```

Code: Scatterplot of the relationship between stolen bases and wins
```{r}
Teams %>% filter(yearID %in% 1961:2001) %>%
    mutate(SB_per_game = SB / G, R_per_game = R / G) %>%
    ggplot(aes(SB_per_game, R_per_game)) + 
    geom_point(alpha = 0.5)

```
Code: Scatterplot of the relationship between bases on balls and runs

```{r}
Teams %>% filter(yearID %in% 1961:2001) %>%
    mutate(BB_per_game = BB / G, R_per_game = R / G) %>%
    ggplot(aes(BB_per_game, R_per_game)) + 
    geom_point(alpha = 0.5)
```

Load the Lahman library. Filter the Teams data frame to include years from 1961 to 2001. Make a scatterplot of runs per game versus at bats (AB) per game.
```{r}
Teams %>% filter(yearID %in% 1961:2001)  %>%
    mutate(AB_per_game = AB / G, R_per_game = R / G) %>%
    ggplot(aes(AB_per_game, R_per_game)) + 
    geom_point(alpha = 0.5)

```

Use the filtered Teams data frame from Question 6. Make a scatterplot of win rate (number of wins per game) versus number of fielding errors (E) per game.
```{r}
Teams %>% filter(yearID %in% 1961:2001)  %>%
    mutate(wins_per_game = W / G, err_per_game =  E/ G) %>%
    ggplot(aes(wins_per_game, err_per_game)) + 
    geom_point(alpha = 0.5)
```

Use the filtered Teams data frame from Question 6. Make a scatterplot of triples (X3B) per game versus doubles (X2B) per game.

Which of the following is true?
```{r}
Teams %>% filter(yearID %in% 1961:2001)  %>%
    mutate(triples_per_game = X3B / G, doubles_per_game =  X2B/ G) %>%
    ggplot(aes(triples_per_game, doubles_per_game)) + 
    geom_point(alpha = 0.5)
```
<h3>Correlation</h3>

Load the Lahman library. Filter the Teams data frame to include years from 1961 to 2001.

What is the correlation coefficient between number of runs per game and number of at bats per game?
```{r}
Teams_small <- Teams %>% filter(yearID %in% 1961:2001)
cor(Teams_small$R/Teams_small$G, Teams_small$AB/Teams_small$G)
```

Use the filtered Teams data frame from Question 7.

What is the correlation coefficient between win rate (number of wins per game) and number of errors per game?

```{r}

cor(Teams_small$W/Teams_small$G, Teams_small$E/Teams_small$G)
```

Use the filtered Teams data frame from Question 7.

What is the correlation coefficient between doubles (X2B) per game and triples (X3B) per game?
```{r}
cor(Teams_small$X2B/Teams_small$G, Teams_small$X3B/Teams_small$G)

```

<h3>Anscombe's Quartet/Stratification
</h3>

```{r}
library(tidyverse)
library(HistData)
data("GaltonFamilies")


set.seed(1983)
galton_heights <- GaltonFamilies %>%
  filter(gender == "male") %>%
  group_by(family) %>%
  sample_n(1) %>%
  ungroup() %>%
  select(father, childHeight) %>%
  rename(son = childHeight)

# number of fathers with height 72 or 72.5 inches
sum(galton_heights$father == 72)
sum(galton_heights$father == 72.5)

# predicted height of a son with a 72 inch tall father
conditional_avg <- galton_heights %>%
    filter(round(father) == 72) %>%
    summarize(avg = mean(son)) %>%
    pull(avg)
conditional_avg

# stratify fathers' heights to make a boxplot of son heights
galton_heights %>% mutate(father_strata = factor(round(father))) %>%
    ggplot(aes(father_strata, son)) +
    geom_boxplot() +
    geom_point()

# center of each boxplot
galton_heights %>%
    mutate(father = round(father)) %>%
    group_by(father) %>%
    summarize(son_conditional_avg = mean(son)) %>%
    ggplot(aes(father, son_conditional_avg)) +
    geom_point()

# calculate values to plot regression line on original data
mu_x <- mean(galton_heights$father)
mu_y <- mean(galton_heights$son)
s_x <- sd(galton_heights$father)
s_y <- sd(galton_heights$son)
r <- cor(galton_heights$father, galton_heights$son)
m <- r * s_y/s_x
b <- mu_y - m*mu_x

# add regression line to plot
galton_heights %>%
    ggplot(aes(father, son)) +
    geom_point(alpha = 0.5) +
    geom_abline(intercept = b, slope = m)
```

<h3>Bivariate Normal Distribution</h3>

When a pair of random variables are approximated by the bivariate normal distribution, scatterplots look like ovals. They can be thin (high correlation) or circle-shaped (no correlation).
When two variables follow a bivariate normal distribution, computing the regression line is equivalent to computing conditional expectations.
We can obtain a much more stable estimate of the conditional expectation by finding the regression line and using it to make predictions.

```{r}
galton_heights %>%
  mutate(z_father = round((father - mean(father)) / sd(father))) %>%
  filter(z_father %in% -2:2) %>%
  ggplot() +  
  stat_qq(aes(sample = son)) +
  facet_wrap( ~ z_father)
```

There are two different regression lines depending on whether we are taking the expectation of Y given X or taking the expectation of X given Y.
```{r}
# compute a regression line to predict the son's height from the father's height
mu_x <- mean(galton_heights$father)
mu_y <- mean(galton_heights$son)
s_x <- sd(galton_heights$father)
s_y <- sd(galton_heights$son)
r <- cor(galton_heights$father, galton_heights$son)
m_1 <-  r * s_y / s_x
b_1 <- mu_y - m_1*mu_x

# compute a regression line to predict the father's height from the son's height
m_2 <-  r * s_x / s_y
b_2 <- mu_x - m_2*mu_y
```



```{r}

set.seed(1989) #if you are using R 3.5 or earlier
set.seed(1989, sample.kind="Rounding") #if you are using R 3.6 or later
library(HistData)
library(tidyverse)
library(dslabs)
data("GaltonFamilies")

female_heights <- GaltonFamilies%>%     
    filter(gender == "female") %>%     
    group_by(family) %>%     
    sample_n(1) %>%     
    ungroup() %>%     
    select(mother, childHeight) %>%     
    rename(daughter = childHeight)

#Daughter's height from mother’s height.
mu_m <- mean(female_heights$mother)
mu_d <- mean(female_heights$daughter)
s_m <- sd(female_heights$mother)
s_d <- sd(female_heights$daughter)

#Correlation coefficient
r <- cor(female_heights$mother, female_heights$daughter)

#Slope of regression line predicting daughters' height from mothers' heights
m_1 <-  r * s_d / s_m

#Intercept of regression line predicting daughters' height from mothers' heights
b_1 <- mu_d - m_1*mu_m

#Change in daughter's height in inches given a 1 inch increase in the mother's height
r * s_y/s_x

#What percent of the variability in daughter heights is explained by the mother's height?
r^2 * 100

```

A mother has a height of 60 inches.

Using the regression formula, what is the conditional expected value of her daughter's height given the mother's height?

```{r}
 m = r * s_y/s_x
b = mu_y - (r * s_y/s_x)*mu_x
x = 60
m*x+b

```

```{r}
# find regression line for predicting runs from BBs
library(tidyverse)
library(Lahman)
bb_slope <- Teams %>% 
  filter(yearID %in% 1961:2001 ) %>% 
  mutate(BB_per_game = BB/G, R_per_game = R/G) %>% 
  lm(R_per_game ~ BB_per_game, data = .) %>% 
  .$coef %>%
  .[2]
bb_slope

# compute regression line for predicting runs from singles
singles_slope <- Teams %>% 
  filter(yearID %in% 1961:2001 ) %>%
  mutate(Singles_per_game = (H-HR-X2B-X3B)/G, R_per_game = R/G) %>%
  lm(R_per_game ~ Singles_per_game, data = .) %>%
  .$coef  %>%
  .[2]
singles_slope

# calculate correlation between HR, BB and singles
Teams %>% 
  filter(yearID %in% 1961:2001 ) %>% 
  mutate(Singles = (H-HR-X2B-X3B)/G, BB = BB/G, HR = HR/G) %>%  
  summarize(cor(BB, HR), cor(Singles, HR), cor(BB,Singles))

```

```{r}
galton_heights <- galton_heights %>%
    mutate(father_centered=father - mean(father))
lm(son ~ father_centered, data = galton_heights)
lm(formula = son ~ father_centered, data = galton_heights)
```

```{r}
# compute RSS for any pair of beta0 and beta1 in Galton's data
library(HistData)
data("GaltonFamilies")
set.seed(1983)
galton_heights <- GaltonFamilies %>%
  filter(gender == "male") %>%
  group_by(family) %>%
  sample_n(1) %>%
  ungroup() %>%
  select(father, childHeight) %>%
  rename(son = childHeight)
rss <- function(beta0, beta1){
    resid <- galton_heights$son - (beta0+beta1*galton_heights$father)
    return(sum(resid^2))
}

# plot RSS as a function of beta1 when beta0=25
beta1 = seq(0, 1, len=nrow(galton_heights))
results <- data.frame(beta1 = beta1,
                      rss = sapply(beta1, rss, beta0 = 25))
results %>% ggplot(aes(beta1, rss)) + geom_line() + 
  geom_line(aes(beta1, rss))

```

```{r}
# fit regression line to predict son's height from father's height
fit <- lm(son ~ father, data = galton_heights)
fit

# summary statistics
summary(fit)
```

LSE are Random Variables

```{r}
# Monte Carlo simulation
B <- 1000
N <- 50
lse <- replicate(B, {
  sample_n(galton_heights, N, replace = TRUE) %>% 
    lm(son ~ father, data = .) %>% 
    .$coef 
})
lse <- data.frame(beta_0 = lse[1,], beta_1 = lse[2,]) 

# Plot the distribution of beta_0 and beta_1
library(gridExtra)
p1 <- lse %>% ggplot(aes(beta_0)) + geom_histogram(binwidth = 5, color = "black") 
p2 <- lse %>% ggplot(aes(beta_1)) + geom_histogram(binwidth = 0.1, color = "black") 
grid.arrange(p1, p2, ncol = 2)

# summary statistics
sample_n(galton_heights, N, replace = TRUE) %>% 
  lm(son ~ father, data = .) %>% 
  summary %>%
  .$coef
  
lse %>% summarize(se_0 = sd(beta_0), se_1 = sd(beta_1))

```

```{r}

lse %>% summarize(cor(beta_0, beta_1))
B <- 1000
N <- 50
lse <- replicate(B, {
      sample_n(galton_heights, N, replace = TRUE) %>%
      mutate(father = father - mean(father)) %>%
      lm(son ~ father, data = .) %>% .$coef 
})
cor(lse[1,], lse[2,]) 
```

The predicted value is often denoted as , which is a random variable. Mathematical theory tells us what the standard error of the predicted value is.
The predict() function in R can give us predictions directly.
```{r}

# plot predictions and confidence intervals
galton_heights %>% ggplot(aes(father, son)) +
  geom_point() +
  geom_smooth(method = "lm")
  
# predict Y directly
fit <- galton_heights %>% lm(son ~ father, data = .) 
Y_hat <- predict(fit, se.fit = TRUE)
names(Y_hat)

# plot best fit line
galton_heights %>%
  mutate(Y_hat = predict(lm(son ~ father, data=.))) %>%
  ggplot(aes(father, Y_hat))+
  geom_line()
```

```{r}

beta1 = seq(0, 1, len=nrow(galton_heights))
results <- data.frame(beta1 = beta1,
                      rss = sapply(beta1, rss, beta0 = 25))
results %>% ggplot(aes(beta1, rss)) + geom_line() + 
  geom_line(aes(beta1, rss), col=2)

```

 In a model for sons’ heights vs fathers’ heights, what is the least squares estimate (LSE) for  if we assume  is 36?
Hint: modify the code above to do your analysis.
```{r}
# compute RSS for any pair of beta0 and beta1 in Galton's data
set.seed(1983)
galton_heights <- GaltonFamilies %>%
  filter(gender == "male") %>%
  group_by(family) %>%
  sample_n(1) %>%
  ungroup() %>%
  select(father, childHeight) %>%
  rename(son = childHeight)
rss <- function(beta0, beta1, data){
    resid <- galton_heights$son - (beta0+beta1*galton_heights$father)
    return(sum(resid^2))
}
# plot RSS as a function of beta1 when beta0=36
beta1 = seq(0, 1, len=nrow(galton_heights))
results <- data.frame(beta1 = beta1,
                      rss = sapply(beta1, rss, beta0 = 36))
results %>% ggplot(aes(beta1, rss)) + geom_line() + 
  geom_line(aes(beta1, rss))
```
 
 
Load the ```Lahman``` library and filter the ```Teams``` data frame to the years 1961-2001.  

Run a linear model in R predicting the number of runs per game based on the number of bases on balls and the number of home runs. Remember to first limit your data to 1961-2001.

What is the coefficient for bases on balls?
```{r}

fit <- Teams %>% filter(yearID %in% 1961:2001) %>% 
mutate(R_per_game = R / G, BB_per_game = BB / G, HR_per_game = HR / G) %>% 
lm(R_per_game ~ BB_per_game + HR_per_game, data = .)
summary(fit)
```

We run a Monte Carlo simulation where we repeatedly take samples of N = 100 from the Galton heights data and compute the regression slope coefficients for each sample:
```{r}

B <- 1000
N <- 100
lse <- replicate(B, {
    sample_n(galton_heights, N, replace = TRUE) %>% 
    lm(son ~ father, data = .) %>% .$coef 
})

lse <- data.frame(beta_0 = lse[1,], beta_1 = lse[2,]) 

```

Which R code(s) below would properly plot the predictions and confidence intervals for our linear model of sons’ heights?

```{r}

 galton_heights %>% ggplot(aes(father, son)) +
    geom_point() +
    geom_smooth(method = "lm")

model <- lm(son ~ father, data = galton_heights)
predictions <- predict(model, interval = c("confidence"), level = 0.95)
data <- as_tibble(predictions) %>% bind_cols(father = galton_heights$father)

ggplot(data, aes(x = father, y = fit)) +
    geom_line(color = "blue", size = 1) + 
    geom_ribbon(aes(ymin=lwr, ymax=upr), alpha=0.2) + 
    geom_point(data = galton_heights, aes(x = father, y = son))
        
```

Define female_heights, a set of mother and daughter heights sampled from GaltonFamilies, as follows:
```{r}
set.seed(1989) #if you are using R 3.5 or earlier
set.seed(1989, sample.kind="Rounding") #if you are using R 3.6 or later
library(HistData)
data("GaltonFamilies")
options(digits = 3)    # report 3 significant digits

female_heights <- GaltonFamilies %>%     
    filter(gender == "female") %>%     
    group_by(family) %>%     
    sample_n(1) %>%     
    ungroup() %>%     
    select(mother, childHeight) %>%     
    rename(daughter = childHeight)

fit <- lm(mother ~ daughter, data = female_heights)
fit$coef[2]
```
Predict mothers' heights using the model.

What is the predicted height of the first mother in the dataset?
```{r}

predict(fit)[1]
```

```{r}
female_heights$mother[1]
```

We have shown how BB and singles have similar predictive power for scoring runs. Another way to compare the usefulness of these baseball metrics is by assessing how stable they are across the years. Because we have to pick players based on their previous performances, we will prefer metrics that are more stable. In these exercises, we will compare the stability of singles and BBs.

Before we get started, we want to generate two tables: one for 2002 and another for the average of 1999-2001 seasons. We want to define per plate appearance statistics, keeping only players with more than 100 plate appearances. Here is how we create the 2002 table:

Now compute a similar table but with rates computed over 1999-2001. Keep only rows from 1999-2001 where players have 100 or more plate appearances, calculate each player's single rate and BB rate per stint (where each row is one stint - a player can have multiple stints within a season), then calculate the average single rate (mean_singles) and average BB rate (mean_bb) per player over the three year period.

How many players had a single rate mean_singles of greater than 0.2 per plate appearance over 1999-2001?


```{r}
library(Lahman)
bat_02 <- Batting %>% filter(yearID == 2002) %>%
    mutate(pa = AB + BB, singles = (H - X2B - X3B - HR)/pa, bb = BB/pa) %>%
    filter(pa >= 100) %>%
    select(playerID, singles, bb)

bat_99_01 <- Batting %>% filter(yearID %in% 1999:2001) %>%
    mutate(pa = AB + BB, singles = (H - X2B - X3B - HR)/pa, bb = BB/pa) %>%
    filter(pa >= 100) %>%
    group_by(playerID) %>%
    summarize(mean_singles = mean(singles), mean_bb = mean(bb))
    
sum(bat_99_01$mean_singles > 0.2)

#How many players had a BB rate ```mean_bb``` of greater than 0.2 per plate appearance over 1999-2001?
sum(bat_99_01$mean_bb > 0.2)
```
Use inner_join() to combine the bat_02 table with the table of 1999-2001 rate averages you created in the previous question.

What is the correlation between 2002 singles rates and 1999-2001 average singles rates?

What is the correlation between 2002 BB rates and 1999-2001 average BB rates?
```{r}

dat <- inner_join(bat_02, bat_99_01)

cor(dat$singles, dat$mean_singles)

cor(dat$bb, dat$mean_bb)
```
Make scatterplots of mean_singles versus singles and mean_bb versus bb.

Are either of these distributions bivariate normal?
```{r}

dat %>%
    ggplot(aes(singles, mean_singles)) +
    geom_point()
dat %>%
    ggplot(aes(bb, mean_bb)) +
    geom_point()
```

12. Fit a linear model to predict 2002 ```singles``` given 1999-2001 ```mean_singles```.

What is the coefficient of ```mean_singles```, the slope of the fit?
```{r}

fit_singles <- lm(singles ~ mean_singles, data = dat)
fit_singles$coef[2]
```


```{r}
# stratify by HR
dat <- Teams %>% filter(yearID %in% 1961:2001) %>%
  mutate(HR = round(HR/G, 1), 
         BB = BB/G,
         R = R/G) %>%
  select(HR, BB, R) %>%
  filter(HR >= 0.4 & HR<=1.2)
  
# calculate slope of regression lines to predict runs by BB in different HR strata
dat %>%  
  group_by(HR) %>%
  summarize(slope = cor(BB,R)*sd(R)/sd(BB))
  
# use lm to get estimated slopes - lm does not work with grouped tibbles
dat %>%  
  group_by(HR) %>%
  lm(R ~ BB, data = .) %>%
  .$coef
  
# inspect a grouped tibble
dat %>% group_by(HR) %>% head()
dat %>% group_by(HR) %>% class()
```


You want to take the tibble dat, which we used in the video on the do() function, and run the linear model R ~ BB for each strata of HR. Then you want to add three new columns to your grouped tibble: the coefficient, standard error, and p-value for the BB term in the model.

What additional code could you write to accomplish your goal?
```{r}

  get_slope <- function(data) {
  fit <- lm(R ~ BB, data = data)
  sum.fit <- summary(fit)

  data.frame(slope = sum.fit$coefficients[2, "Estimate"], 
             se = sum.fit$coefficients[2, "Std. Error"],
             pvalue = sum.fit$coefficients[2, "Pr(>|t|)"])
  }
          dat %>% 
  group_by(HR) %>% 
  do(get_slope(.))
```

You want to know whether the relationship between home runs and runs per game varies by baseball league. You create the following dataset:

```{r}
dat %>% 
  group_by(lgID) %>% 
  do(tidy(lm(R ~ HR, data = .), conf.int = T)) %>% 
  filter(term == "HR") 
```

We have investigated the relationship between fathers' heights and sons' heights. But what about other parent-child relationships? Does one parent's height have a stronger association with child height? How does the child's gender affect this relationship in heights? Are any differences that we observe statistically significant?

The galton dataset is a sample of one male and one female child from each family in the GaltonFamilies dataset. The pair column denotes whether the pair is father and daughter, father and son, mother and daughter, or mother and son.

Create the galton dataset using the code below:

```{r}
library(tidyverse)
library(HistData)
data("GaltonFamilies")
set.seed(1) # if you are using R 3.5 or earlier
set.seed(1, sample.kind = "Rounding") # if you are using R 3.6 or later
galton <- GaltonFamilies %>%
    group_by(family, gender) %>%
    sample_n(1) %>%
    ungroup() %>% 
    gather(parent, parentHeight, father:mother) %>%
    mutate(child = ifelse(gender == "female", "daughter", "son")) %>%
    unite(pair, c("parent", "child"))

galton
```
Group by pair and summarize the number of observations in each group.

How many father-daughter pairs are in the dataset?
```{r}
summary_group <- galton %>% group_by(pair) %>%
    summarise(Count = n())   
```

Calculate the correlation coefficients for fathers and daughters, fathers and sons, mothers and daughters and mothers and sons.

Which pair has the strongest correlation in heights?

```{r}
galton %>%
    group_by(pair) %>%
    summarize(cor = cor(parentHeight, childHeight)) %>%
    filter(cor == max(cor))
```

Use lm() and the broom package to fit regression lines for each parent-child pair type. Compute the least squares estimates, standard errors, confidence intervals and p-values for the parentHeight coefficient for each pair.

What is the estimate of the father-daughter coefficient?

```{r}
library('broom')

galton %>%
    group_by(pair) %>%
    do(tidy(lm(childHeight ~ parentHeight, data = .), conf.int = TRUE)) %>%
    filter(term == "parentHeight", pair == "father_daughter") %>%
    pull(estimate)
```

For every 1-inch increase in mother's height, how many inches does the typical son's height increase?
```{r}

galton %>% group_by(pair) %>% 
          do(tidy(lm(childHeight ~ parentHeight,data = .),conf.int=TRUE)) %>%
          filter(term == "parentHeight", pair == "mother_son") %>%
    pull(estimate)
```

10b. Which sets of parent-child heights are significantly correlated at a p-value cut off of .05?

```{r}
galton %>%
    group_by(pair) %>%
    do(tidy(lm(childHeight ~ parentHeight, data = .), conf.int = TRUE)) %>%
    filter(term == "parentHeight" & p.value < .05)
```

```{r}

library(reshape2)
library(lpSolve)

players <- players %>% filter(debut <= "1997-01-01" & debut > "1988-01-01")
constraint_matrix <- acast(players, POS ~ playerID, fun.aggregate = length)
npos <- nrow(constraint_matrix)
constraint_matrix <- rbind(constraint_matrix, salary = players$salary)
constraint_dir <- c(rep("==", npos), "<=")
constraint_limit <- c(rep(1, npos), 50*10^6)
lp_solution <- lp("max", players$R_hat,
                  constraint_matrix, constraint_dir, constraint_limit,
                  all.int = TRUE) 
```
This algorithm chooses these 9 players:

```{r}
our_team <- players %>%
  filter(lp_solution$solution == 1) %>%
  arrange(desc(R_hat))
our_team %>% select(nameFirst, nameLast, POS, salary, R_hat)
```
We note that these players all have above average BB and HR rates while the same is not true for singles.

```{r}
my_scale <- function(x) (x - median(x))/mad(x)
players %>% mutate(BB = my_scale(BB), 
                   singles = my_scale(singles),
                   doubles = my_scale(doubles),
                   triples = my_scale(triples),
                   HR = my_scale(HR),
                   AVG = my_scale(AVG),
                   R_hat = my_scale(R_hat)) %>%
    filter(playerID %in% our_team$playerID) %>%
    select(nameFirst, nameLast, BB, singles, doubles, triples, HR, AVG, R_hat) %>%
    arrange(desc(R_hat))
```

